---
title: "Global Flight Route Density Mapping"
author: "Enhanced R Tutorial"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: show
    code-tools: true
    theme: cosmo
    fig-width: 16
    fig-height: 10
    embed-resources: true
execute:
  warning: false
  message: false
---

## Overview

This tutorial guides you through creating professional visualizations of global flight routes using R. You'll learn how to transform straight lines into **curved routes that follow Earth's surface** (similar to Python's Cartopy `transform=ccrs.Geodetic()`), create density heatmaps, and analyze aviation patterns worldwide.

### What You'll Create

1. **Airport location maps** - See where airports cluster globally
2. **Straight flight routes** - Direct lines between airports
3. **Geodetic flight routes** - Realistic curved paths following Earth's curvature
4. **Hexagonal density analysis** - Count routes passing through each region
5. **Route density heatmaps** - Visualize aviation traffic hot spots
6. **Combined visualizations** - Overlay routes on density maps

### Key Concepts Explained

**Robinson Projection**: A map projection that balances shape and area distortion, making the world look more realistic than flat projections. Perfect for global datasets.

**Geodetic Lines**: Lines that follow the shortest path on Earth's curved surface (great circle routes). When you fly from New York to Tokyo, planes don't follow a straight line on a flat map—they curve northward over Alaska because Earth is a sphere.

**Hexagonal Grids**: A honeycomb pattern that divides the world into equal-sized cells for counting and analyzing spatial patterns.

## Prerequisites

### Required R Packages

```{r}
#| eval: false
install.packages(c("readr", "dplyr", "stringr", "sf", "ggplot2", 
                   "viridis", "scales", "rnaturalearth"))
```

### Load Libraries

```{r}
library(readr)      # Reading CSV files
library(dplyr)      # Data manipulation
library(stringr)    # String operations
library(sf)         # Spatial data handling
library(ggplot2)    # Plotting
library(viridis)    # Color scales
library(scales)     # Number formatting
library(rnaturalearth)  # World map data
```

## Step 1: Download and Prepare Data

### Understanding the Data

We'll use **OpenFlights data**, a free database of airports and flight routes last updated in 2014. While not current, it provides excellent data for learning visualization techniques.

- **airports.dat**: Contains ~7,000 airports with coordinates
- **routes.dat**: Contains ~67,000 flight routes between airports

```{r}
# Create folders for storing data and output files
folders <- c("output", "data")
sapply(folders, dir.create, showWarnings = FALSE, recursive = TRUE)

# Download airports data if not already present
if (!file.exists("data/airports.dat")) {
  cat("Downloading airports.dat to the data folder...\n")
  download.file(
    "https://raw.githubusercontent.com/jpatokal/openflights/master/data/airports.dat",
    "data/airports.dat",
    mode = "wb"
  )
  cat("✓ Airports data downloaded\n")
}

# Download routes data if not already present
if (!file.exists("data/routes.dat")) {
  cat("Downloading routes.dat to the data folder...\n")
  download.file(
    "https://raw.githubusercontent.com/jpatokal/openflights/master/data/routes.dat",
    "data/routes.dat",
    mode = "wb"
  )
  cat("✓ Routes data downloaded\n")
}
```

### Load Routes Data

```{r}
# Define column names for the routes dataset
routes_column <- c(
    "Airline",                  # Airline code
    "Airline_ID",               # Unique airline ID
    "Source_airport",           # Origin airport (3-letter IATA code)
    "Source_airport_ID",        # Origin airport ID
    "Destination_airport",      # Destination airport (3-letter IATA code)
    "Destination_airport_ID",   # Destination airport ID
    "Codeshare",                # Is this a codeshare flight?
    "Stops",                    # Number of stops
    "Equipment"                 # Aircraft types used
)

# Load the routes data
routes <- read_delim("data/routes.dat", delim = ",", 
                     col_names = routes_column, 
                     show_col_types = FALSE)

cat("✓ Loaded", format(nrow(routes), big.mark = ","), "routes\n")
```

### Load and Process Airports Data

```{r}
# Define column names for the airports dataset
airports_column <- c(
    "Airport_ID",           # Unique airport identifier
    "Name",                 # Airport name
    "City",                 # City name
    "Country",              # Country name
    "IATA",                 # 3-letter IATA code (e.g., JFK, LHR)
    "ICAO",                 # 4-letter ICAO code
    "Latitude",             # Latitude coordinate
    "Longitude",            # Longitude coordinate
    "Altitude",             # Altitude in feet
    "Timezone",             # Hours offset from UTC
    "DST",                  # Daylight savings time
    "Tz_database_timezone", # Timezone database name
    "Type",                 # Airport type
    "Source"                # Data source
)

# Load the airports data
airports <- read_delim("data/airports.dat", delim = ",", 
                       col_names = airports_column, 
                       show_col_types = FALSE)

cat("✓ Loaded", format(nrow(airports), big.mark = ","), "airports\n")
```

### Convert Airports to Spatial Format

This step transforms our simple latitude/longitude table into a **spatial object** that R can use for mapping and analysis.

```{r}
# Convert airports dataframe to spatial features (sf) object
airports_sf <- airports %>%
  mutate(
    Latitude = as.numeric(Latitude), 
    Longitude = as.numeric(Longitude)
  ) %>%
  filter(!is.na(Latitude), !is.na(Longitude)) %>%
  # Create point geometries from coordinates
  st_as_sf(coords = c("Longitude", "Latitude"), 
           crs = 4326,  # WGS84 coordinate system
           remove = FALSE)

# Transform to Robinson projection for realistic global view
airports_sf <- st_transform(airports_sf, crs = "+proj=robin")

cat("✓ Created spatial object with", 
    format(nrow(airports_sf), big.mark = ","), 
    "airports\n")
```

::: {.callout-note}
## Understanding Coordinate Reference Systems (CRS)

- **CRS 4326 (WGS84)**: Standard latitude/longitude coordinates (what GPS uses)
- **Robinson Projection (`+proj=robin`)**: A balanced world map projection that minimizes distortion across the globe. It's more realistic than flat projections because Earth is round, not flat.

Think of it like trying to peel an orange and flatten the peel—you'll always get some distortion. Robinson projection spreads that distortion evenly.
:::

## Step 2: Create Geodetic Lines (Curved Routes)

### Why Curved Lines Matter

On a flat map, the shortest distance looks like a straight line. But Earth is a sphere! The actual shortest distance between two points on a sphere is a **great circle route**—a curve that follows Earth's surface.

**Example**: A flight from New York to Tokyo appears to curve northward over Alaska on a map. This isn't a detour—it's actually the shortest path on Earth's curved surface!

### The R Equivalent of Cartopy's `transform=ccrs.Geodetic()`

In Python's Cartopy, you use `transform=ccrs.Geodetic()` to create curved lines. In R, we achieve the same effect using `st_segmentize()`, which adds intermediate points along the line to follow Earth's curvature.

### Clean Routes Data

First, we remove any invalid route entries:

```{r}
# Clean routes: keep only valid 3-letter IATA codes
routes_clean <- routes %>%
  mutate(
    Source_airport = str_trim(Source_airport),
    Destination_airport = str_trim(Destination_airport)
  ) %>%
  filter(
    !is.na(Source_airport), !is.na(Destination_airport),
    Source_airport != "", Destination_airport != "",
    # IATA codes are exactly 3 capital letters (e.g., JFK, LHR, SIN)
    str_detect(Source_airport, "^[A-Z]{3}$"),
    str_detect(Destination_airport, "^[A-Z]{3}$")
  )

# Report cleaning results
removed <- nrow(routes) - nrow(routes_clean)
if (removed > 0) {
  cat("⚠ Removed", format(removed, big.mark = ","), 
      "invalid entries\n")
}
cat("✓ Kept", format(nrow(routes_clean), big.mark = ","), 
    "valid routes\n")
```

### Match Routes with Airport Coordinates

Now we need to find the latitude/longitude for each airport:

```{r}
# Create a lookup table: IATA code → coordinates
airports_lookup <- airports %>%
  mutate(
    IATA = str_trim(IATA),
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude)
  ) %>%
  filter(
    !is.na(IATA), IATA != "",
    str_detect(IATA, "^[A-Z]{3}$"),
    !is.na(Latitude), !is.na(Longitude)
  ) %>%
  distinct(IATA, .keep_all = TRUE) %>%
  select(IATA, Latitude, Longitude)

# Join routes with coordinates for both origin and destination
routes_joined <- routes_clean %>%
  left_join(
    airports_lookup %>% 
      rename(src_lat = Latitude, src_lon = Longitude),
    by = c("Source_airport" = "IATA")
  ) %>%
  left_join(
    airports_lookup %>% 
      rename(dst_lat = Latitude, dst_lon = Longitude),
    by = c("Destination_airport" = "IATA")
  ) %>%
  filter(
    !is.na(src_lon), !is.na(src_lat),
    !is.na(dst_lon), !is.na(dst_lat)
  )

cat("✓ Matched", format(nrow(routes_joined), big.mark = ","), 
    "routes with coordinates\n")
```

### Create Geodetic Line Geometries

Here's the **key step** that creates curved lines following Earth's surface:

```{r}
# Create curved lines that follow Earth's curvature
# This is the R equivalent of Cartopy's transform=ccrs.Geodetic()
routes_lines_sf <- routes_joined %>%
  mutate(
    geometry = mapply(
      function(x1, y1, x2, y2) {
        # Step 1: Create a basic straight line
        line <- st_linestring(
          matrix(c(x1, y1, x2, y2), ncol = 2, byrow = TRUE)
        )
        # Step 2: Add points every 100km to follow Earth's curve
        # This transforms the straight line into a geodetic (curved) path
        st_segmentize(line, dfMaxLength = 100000)  # 100km intervals
      },
      src_lon, src_lat, dst_lon, dst_lat,
      SIMPLIFY = FALSE
    ) %>% st_sfc(crs = 4326)  # WGS84 coordinate system
  ) %>%
  st_as_sf()

# Transform to Robinson projection
routes_lines_sf <- st_transform(routes_lines_sf, crs = "+proj=robin")

cat("✓ Created", format(nrow(routes_lines_sf), big.mark = ","), 
    "geodetic route lines\n")
```

::: {.callout-tip}
## How `st_segmentize()` Creates Curved Lines

Think of it like drawing a curve with a ruler:

1. **Without segmentization**: A straight line from point A to B (incorrect on a sphere)
2. **With segmentization**: Adds points every 100km, then connects them with short segments that follow Earth's surface (correct geodetic path)

The `dfMaxLength = 100000` parameter means "add a point every 100 kilometers." Smaller values = smoother curves but slower processing.
:::

## Step 3: Create Hexagonal Grid for Density Analysis

### Why Hexagons?

Hexagons are ideal for spatial analysis because:

- They have **uniform adjacency** (each hexagon touches 6 neighbors equally)
- They're more **circular** than squares (better represent areas around a point)
- They look professional and modern

We'll use these hexagons to count how many routes pass through each region.

```{r}
# Load world boundaries with Robinson projection
world <- ne_countries(scale = "small", returnclass = "sf") %>%
  st_transform("+proj=robin") %>%
  st_make_valid()

# Create hexagonal grid over the world
# cellsize = 100,000 meters (100km) means each hexagon is ~100km wide
hexgrid <- st_make_grid(
  world, 
  cellsize = 1e5,      # 100km hexagons
  what = "polygons", 
  square = FALSE       # FALSE = hexagons, TRUE = squares
) %>%
  st_as_sf() %>%
  mutate(id = row_number())

# Keep only hexagons that touch land (remove ocean-only hexagons)
land_union <- st_union(world)
hexgrid_world <- hexgrid[
  st_intersects(hexgrid, land_union, sparse = FALSE), 
]

cat("✓ Created", format(nrow(hexgrid_world), big.mark = ","), 
    "hexagons (~100km width each)\n")

# Preview the grid
plot(st_geometry(world), col = "lightgray", border = "white", 
     main = "Hexagonal Grid Preview", cex.main = 1.5)
plot(st_geometry(hexgrid_world), add = TRUE, 
     border = "red", col = NA, lwd = 0.3)
```

::: {.callout-tip}
## Adjusting Hexagon Size

Change the `cellsize` parameter to adjust detail level:

- `cellsize = 5e4` → 50km hexagons (more detail, slower)
- `cellsize = 1e5` → 100km hexagons (balanced, **recommended**)
- `cellsize = 2e5` → 200km hexagons (faster, less detail)

Smaller hexagons = more precise analysis but longer processing time.
:::

## Step 4: Count Routes per Hexagon

Now we'll count how many **curved (geodetic) routes** pass through each hexagon. This is crucial—we're using the realistic curved lines, not straight lines!

```{r}
# Spatial intersection: which routes pass through which hexagons?
# This counts where the CURVED routes cross each hexagon
intersections <- st_intersects(hexgrid_world, routes_lines_sf)

# Add route count to each hexagon
hexgrid_world_rt_cnt <- hexgrid_world %>%
  mutate(nm_rts = lengths(intersections))

# Show statistics
cat("\n📊 Route Density Statistics:\n")
cat("─────────────────────────────\n")
summary_stats <- summary(hexgrid_world_rt_cnt$nm_rts)
print(summary_stats)
cat("\nHighest density hexagon:", 
    format(max(hexgrid_world_rt_cnt$nm_rts), big.mark = ","), 
    "routes\n")
```

::: {.callout-note}
## What This Analysis Shows

Each hexagon's count represents the **number of flight routes passing through that region**. High counts indicate:

- Major flight corridors (e.g., North Atlantic)
- Hub airports (e.g., Dubai, Atlanta, London)
- Dense regional networks (e.g., Western Europe, East Coast USA)

Low counts indicate remote areas with limited air connectivity.
:::

## Step 5: Visualizations

### Visualization 1: Airport Locations

First, let's see where airports are distributed globally:

```{r}
#| label: fig-airports
#| fig-cap: "Global distribution of airports showing concentration in developed regions"

map_airports <- ggplot() +
  geom_sf(data = world, fill = "gray20", color = "gray40", 
          size = 0.1) +
  geom_sf(data = airports_sf, color = "#FFD700", 
          size = 0.3, alpha = 0.6) +
  labs(
    title = "Global Airport Locations",
    subtitle = paste(format(nrow(airports_sf), big.mark = ","), 
                     "airports worldwide"),
    caption = "Source: jpatokal/openflights (2014) | Projection: Robinson"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 18, face = "bold", 
                              hjust = 0.5, color = "white"),
    plot.subtitle = element_text(size = 12, hjust = 0.5, 
                                  color = "gray70"),
    plot.caption = element_text(size = 8, color = "gray50", 
                                 hjust = 1),
    plot.background = element_rect(fill = "gray10", color = NA),
    panel.background = element_rect(fill = "gray10", color = NA)
  ) +
  coord_sf(crs = "+proj=robin")

print(map_airports)
```

**What to Look For**: Notice how airports cluster in North America, Europe, and East Asia. Sparse coverage in central Africa and South America.

### Visualization 2: Flight Routes (Straight Lines)

This shows routes as simple straight lines—**not geographically accurate**, but useful for comparison:

```{r}
#| label: fig-routes-straight
#| fig-cap: "Flight routes as straight lines (not following Earth's curvature)"

# Sample 5,000 routes for better visibility
set.seed(123)
routes_sample <- routes_lines_sf %>% 
  slice_sample(n = min(5000, nrow(routes_lines_sf)))

map_routes_straight <- ggplot() +
  geom_sf(data = world, fill = "gray20", color = "gray40", 
          size = 0.1) +
  geom_sf(data = routes_sample, color = "#00BFFF", 
          size = 0.1, alpha = 0.05) +
  labs(
    title = "Global Flight Routes (Straight Lines - Not Realistic)",
    subtitle = paste("Showing", 
                     format(nrow(routes_sample), big.mark = ","), 
                     "sampled routes"),
    caption = "Source: jpatokal/openflights (2014) | Projection: Robinson"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 18, face = "bold", 
                              hjust = 0.5, color = "white"),
    plot.subtitle = element_text(size = 12, hjust = 0.5, 
                                  color = "gray70"),
    plot.caption = element_text(size = 8, color = "gray50", 
                                 hjust = 1),
    plot.background = element_rect(fill = "gray10", color = NA),
    panel.background = element_rect(fill = "gray10", color = NA)
  ) +
  coord_sf(crs = "+proj=robin")

print(map_routes_straight)
```

### Visualization 3: Geodetic Routes (Following Earth's Curvature)

Now see the **realistic curved routes** that follow Earth's surface:

```{r}
#| label: fig-routes-geodetic
#| fig-cap: "Flight routes following great circle paths (geodetic) - geographically accurate"

map_routes_curved <- ggplot() +
  geom_sf(data = world, fill = "gray20", color = "gray40", 
          size = 0.1) +
  geom_sf(data = routes_sample, color = "#FF6B6B", 
          size = 0.1, alpha = 0.1) +  # Increased to 0.1 for clarity
  labs(
    title = "Global Flight Routes (Geodetic - Following Earth's Curvature)",
    subtitle = paste("Showing", 
                     format(nrow(routes_sample), big.mark = ","), 
                     "routes as great circle paths"),
    caption = "Source: jpatokal/openflights (2014) | Projection: Robinson"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 18, face = "bold", 
                              hjust = 0.5, color = "white"),
    plot.subtitle = element_text(size = 12, hjust = 0.5, 
                                  color = "gray70"),
    plot.caption = element_text(size = 8, color = "gray50", 
                                 hjust = 1),
    plot.background = element_rect(fill = "gray10", color = NA),
    panel.background = element_rect(fill = "gray10", color = NA)
  ) +
  coord_sf(crs = "+proj=robin")

print(map_routes_curved)
```

::: {.callout-important}
## Compare Straight vs Curved Routes!

Look at transoceanic routes (e.g., USA to Europe, USA to Asia):

- **Straight lines** appear to go directly across
- **Geodetic lines** curve northward (over Greenland, Alaska) because that's the **actual shortest path** on Earth's sphere

Airlines use geodetic routes to save fuel and time!
:::

### Visualization 4: Hexagonal Grid Overlay

See the analysis grid we created:

```{r}
#| label: fig-hexgrid
#| fig-cap: "Hexagonal analysis grid over world map - each cell is ~100km wide"

map_hexgrid <- ggplot() +
  geom_sf(data = world, fill = "gray95", color = "white", 
          size = 0.1) +
  geom_sf(data = hexgrid_world, fill = NA, color = "red", 
          size = 0.2, alpha = 0.5) +
  labs(
    title = "Hexagonal Grid Over World Map",
    subtitle = paste(format(nrow(hexgrid_world), big.mark = ","), 
                     "hexagons (~100km width each)"),
    caption = "Source: Natural Earth | Projection: Robinson"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, 
                                  color = "gray30"),
    plot.caption = element_text(size = 8, color = "gray50", 
                                 hjust = 1),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  ) +
  coord_sf(crs = "+proj=robin")

print(map_hexgrid)
```

### Visualization 5: Flight Route Density Heatmap

The main analysis—see where air traffic is densest:

```{r}
#| label: fig-density
#| fig-cap: "Flight route density heatmap showing aviation traffic hot spots"

map_density <- ggplot() +
  geom_sf(data = world, fill = "gray95", color = "white", 
          size = 0.1) +
  geom_sf(data = hexgrid_world_rt_cnt, 
          aes(fill = nm_rts), 
          color = NA,
          alpha = 0.85) +
  scale_fill_viridis_c(
    option = "plasma",
    name = "Number of\nFlight Routes",
    trans = "sqrt",  # Square root scale for better visualization
    breaks = c(0, 10, 50, 100, 250, 500, 1000, 2500, 5000),
    labels = comma,
    guide = guide_colorbar(
      barwidth = 15,
      barheight = 0.5,
      title.position = "top",
      title.hjust = 0.5
    )
  ) +
  labs(
    title = "Global Flight Route Density (Based on Geodetic Routes)",
    subtitle = "Hexagonal grid showing number of curved flight routes intersecting each cell",
    caption = "Source: jpatokal/openflights (2014) | Projection: Robinson"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, 
                                  color = "gray30"),
    plot.caption = element_text(size = 8, color = "gray50", 
                                 hjust = 1),
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.title = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  ) +
  coord_sf(crs = "+proj=robin")

print(map_density)
```

::: {.callout-note}
## Interpreting the Density Map

**Hot spots (yellow/white):**

- **Eastern USA**: New York, Atlanta, Chicago hubs
- **Western Europe**: London, Paris, Frankfurt network
- **East Asia**: Tokyo, Shanghai, Hong Kong, Singapore
- **Middle East**: Dubai and Doha mega-hubs

**Cold spots (dark purple):**

- Central Africa
- Central Asia
- Amazon rainforest
- Sahara Desert
- Remote oceanic regions

The square root transformation (`trans = "sqrt"`) compresses high values so you can see both busy and quiet areas clearly.
:::

### Visualization 6: Combined Routes and Density

The ultimate visualization combining individual routes with the density heatmap:

```{r}
#| label: fig-combined
#| fig-cap: "Combined visualization showing both individual flight routes and regional density"

map_combined <- ggplot() +
  geom_sf(data = world, fill = "gray20", color = "gray40", 
          size = 0.1) +
  geom_sf(data = hexgrid_world_rt_cnt, 
          aes(fill = nm_rts), 
          color = NA,
          alpha = 0.7) +
  geom_sf(data = routes_sample, color = "white", 
          size = 0.05, alpha = 0.02) +
  scale_fill_viridis_c(
    option = "inferno",
    name = "Route\nDensity",
    trans = "sqrt",
    breaks = c(0, 50, 250, 1000, 2500),
    labels = comma,
    guide = guide_colorbar(
      barwidth = 12,
      barheight = 0.5,
      title.position = "top",
      title.hjust = 0.5
    )
  ) +
  labs(
    title = "Global Aviation Network: Geodetic Routes and Density",
    subtitle = "Combining individual curved flight paths with regional density analysis",
    caption = "Source: jpatokal/openflights (2014) | Projection: Robinson"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 18, face = "bold", 
                              hjust = 0.5, color = "white"),
    plot.subtitle = element_text(size = 12, hjust = 0.5, 
                                  color = "gray70"),
    plot.caption = element_text(size = 8, color = "gray50", 
                                 hjust = 1),
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold", 
                                color = "white"),
    legend.text = element_text(size = 8, color = "gray70"),
    plot.background = element_rect(fill = "gray10", color = NA),
    panel.background = element_rect(fill = "gray10", color = NA)
  ) +
  coord_sf(crs = "+proj=robin")

print(map_combined)
```

## Step 6: Save High-Quality Outputs

Export all your beautiful maps:

```{r}
#| label: save-outputs

# Create output folder if it doesn't exist
dir.create("output", showWarnings = FALSE)

# Save all visualizations as high-resolution PNG files
ggsave("output/01_airport_locations.png", 
       plot = map_airports, 
       width = 14, height = 9, dpi = 300, bg = "gray10")

ggsave("output/02_routes_straight.png", 
       plot = map_routes_straight, 
       width = 14, height = 9, dpi = 300, bg = "gray10")

ggsave("output/03_routes_curved_geodetic.png", 
       plot = map_routes_curved, 
       width = 14, height = 9, dpi = 300, bg = "gray10")

ggsave("output/04_hexgrid_overlay.png", 
       plot = map_hexgrid, 
       width = 14, height = 9, dpi = 300, bg = "white")

ggsave("output/05_flight_density_heatmap.png", 
       plot = map_density, 
       width = 14, height = 9, dpi = 300, bg = "white")

# Also save density map as PDF (vector format, infinite zoom!)
ggsave("output/05_flight_density_heatmap.pdf", 
       plot = map_density, 
       width = 14, height = 9, device = "pdf")

ggsave("output/06_combined_routes_density.png", 
       plot = map_combined, 
       width = 14, height = 9, dpi = 300, bg = "gray10")

cat("✅ All visualizations saved to output folder!\n")
cat("   Check the 'output' directory for your maps.\n")
```
